---
title: "Spatial Analysis in R, Lab"
author: "Environmental Data Analytics | John Fay & Luana Lima"
date: "Fall 2025"
output: pdf_document
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: none
---

## Topics
-   Q&A on Lessons
    -   Importance of spatial data in environmental data analysis
    -   Spatial data models: Raster v. Vector
    -   Spatial data file formats; reading in spatial data
    -   `sf` objects, aka "the spatial dataframe" & its components
    -   Coordinate reference systems
    -   Attribute joins
    -   Data aggregation; dissolving features
    -   Spatial selection
    -   Spatial intersection
    -   Geometric modification (centroid, buffer, convex hull, etc.)
    -   Data Visualization
-   Demo: Spatial Data Resources
    - Open Data Hubs
    - ArcGIS Online Data
    
    
## Spatal Data Resources

### Background
-   More and more organizations provide "Open Data Portals"
-   The "Manufactured gas plant" dataset we used was obtained via one of these.
-   Let's look at the process of exploring and ingesting these data into R
-   First, let's set up our coding session

```{r setup, warning=F, error=F}
#Import packages
library(tidyverse);library(here);
library(sf);library(mapview);library(leaflet)

#A new package: arcgis -- discussed below
#install.packages('arcgis', repos = c('https://r-arcgis.r-universe.dev', 'https://cloud.r-project.org'))
#library(arcgis)

#Check the working directory
here()

```


### Exercise: Open Data Portals
*   Navigate to https://www.nconemap.gov/ - NC's home for spatial data
  -   Scroll down and view the various themes you can browse...
  -   Click on the Energy theme...
  -   Note the "Manufactured Gas Plant Site" layer, listed as a *Feature Service*
  
*   Click on the link for the "Manufactured Gas Plant Site" layer
  -   Direct link: https://www.nconemap.gov/datasets/a0722062315143a7980cebb79a204781_4/

*   From the above website, click on the`View Full Details` link. 
  -   From here you could download the data, but we'll link to the data directly. 
  -   Find and extend the dropdown list for `View API Resources`
  -   Copy the link associated with the `GeoJSON` service
  
The link you should have copied is here:  
https://services.nconemap.gov/secure/rest/services/NC1Map_Hazardous_Waste/MapServer/4/query?outFields=*&where=1%3D1&f=geojson 

We'll now use this to fetch the data directly into our R coding environment.

```{r import data from GeoJSON feature service}
#Set the URL as the data source string
data_source <- paste0(
  'https://services.nconemap.gov/secure/rest/services/',
  'NC1Map_Hazardous_Waste/MapServer/4/',
  'query?outFields=*&where=1%3D1&f=geojson')

gas_plants <- st_read(data_source)

mapview(gas_plants)

```


Another resource is hosted by the US Fish & Wildlife Their open data site (<https://gis-fws.opendata.arcgis.com/>) has links to many datasets.

When the data is served in GeoJSON format, we can ingest it directly in to R. Follow these steps: 
-   Navigate to <https://gis-fws.opendata.arcgis.com/>
-   Scroll down to the Explore Categories area. Select Biota (for example)
-   Search for "Sea Turtle Nesting"
-   Select the first return ("2023 Sea Turtle Nesting Survey TAFB results") and open its [link](https://duke.is/SeaTurtleNestingData)
-   Locate the APIs dropdown and copy the link to the GeoJSON option.

If you have difficulty, the link you want is: 
-   <https://services.arcgis.com/QVENGdaPbd4LUkLV/arcgis/rest/services/survey123_3d7e213f47a94f1b829aa7cf54eab095_results/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson>

This a link to a spatial dataset in GeoJSON format. Let's read this dataset in and explore it:
-   Create a spatial dataframe from the link, filtering for only records in North Carolina (STATE == "NC")
-   Reveal its structure
-   What coordinate reference system (crs) does it use?

Well do this one together...
```{r read in sea turtle data}
#Create the data source url
data_source1 <- paste0(
  'https://services.arcgis.com/QVENGdaPbd4LUkLV/arcgis/rest/',
  'services/survey123_3d7e213f47a94f1b829aa7cf54eab095_results/',
  'FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson')

#Read in to a spatial dataframe
sea_turtles <- st_read(data_source1)

mapview(sea_turtles)

```

#Now we can make a pretty map!
```{r cluster map}
#Create the map
turtle_map <- leaflet(sea_turtles) %>% 
  addTiles() %>% 
  addMarkers(
    clusterOptions = markerClusterOptions(),  #cluster option 
    popup = ~species
  )

#Show the map
turtle_map

#Saved the map as an HTML file
#htmlwidgets::saveWidget(turtle_map, 'my_map.html')
```


Now, see if you can find, import, and display data from Durham's Open Data portal
-   Do a web search for Durham Open Data
-   Browse and choose a Feature Service (e.g. Landmarks & Points of Interest)
-   Find the GeoJSON endpoint and import the dataset
-   Explore properties of the dataset
-   Make a map
-   Export your map to an image or HTML document


```{r fetch and display an online spatial dataset}


```



## ArcGIS Online
If we dissect the URL pointing to the feature server, we can discover other data hosted on the server. Both the gas plant and the landslide data sources begin with <"https://services.nconemap.gov/secure/rest/services/"> -- the address of the server hosting these layers. 
-   Open <https://services.nconemap.gov/secure/rest/services/> in your browser
-   Note all the resources hosted there
  -   **MapServer** and **FeatureServer** items host vector data
-   Select the **NC1Map_Health** Feature Server
  -   View the metadata for this feature service
  -   This feature service hosts four layers...
-   Select the Hospitals (0)  Layer link
  -   We see more metadata, including a list of all the fields in this spatial dataset
  -   At the bottom, we see "Supported Operations" including a Query link
-   Select the Query link at the bottom of the page
  

Recognizing that R users have a thirst for spatial data hosted on ArcGIS Servers, ESRI has developed a tool that allows easier access to hosted spatial data. Info on the ArcGIS Package for R is here:   
https://www.esri.com/arcgis-blog/products/developers/announcements/announcing-arcgis-r-package/

We can use this package to access the same data we found above: 

```{r eval=FALSE}
#The same data source for the manufactured gas plant data
f_service_url <- paste0('https://services.nconemap.gov/secure/rest/services/NC1Map_Health/FeatureServer')

#Read the data into a feature service object
f_service <- arc_open(f_service_url)
class(f_service)

#Retrieve the data associated with the 1st feature service w/`get_layer()` 
f_layer <- get_layer(f_service, id=0)
class(f_layer)

#Convert the feature layer to a spatial dataframe w/`arc_select()`
hospitals_sf <- arc_select(f_layer)
class(hospitals_sf)

#Map the spatial dataframe
mapview::mapview(
  hospitals_sf,
  zcol='hltype',
  col.regions = RColorBrewer::brewer.pal(10, "Paired"),
  map.types = "CartoDB.Positron"
  )

```
