---
title: "Assignment 10: Data Scraping"
author: "Sarah Hall"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
library(lubridate)
library(rvest)
library(here)
here()

#create my theme 
mytheme <- theme_classic(base_size = 12) 
  theme(axis.text = element_text(color = "black"),
        legend.position = "top")
#set mytheme to standard theme
theme_set(mytheme)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
LWSP_website <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024')

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
water_system_name <- LWSP_website %>% 
  html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>% 
  html_text()
PWSID <- LWSP_website %>%  
  html_nodes('td tr:nth-child(1) td:nth-child(5)') %>% 
  html_text()
Ownership <- LWSP_website %>%  
  html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>%
  html_text()
MaxDayUse <- LWSP_website %>%  
  html_nodes('th~ td+ td , th~ td+ td') %>%
  html_text()

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 
water_df <- data.frame("Month" = c(1,5,9,2,6,10,3,7,11,4,8,12),
                       "Year" = rep(2024, 12),
                       "MaxDayUse" = as.numeric(MaxDayUse)) %>% 
            mutate(
              Water_System_Name = !!Ownership,
              PWSID = !!PWSID,
              Ownership = !!Ownership,
              Date = my(paste(Month, "-", Year)))

#5 
ggplot(water_df, aes(x=Date,
                     y = MaxDayUse))+
  geom_line()+
  labs(
    title = "2024 Durham Muncipal Water Supply Maximum Daily Withdrawal",
    y = "Maximum Daily Withdrawal (mgd)",
    x = "Date")

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs - "PWSID" and "year" - that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. 

scrape.it <- function(the_facility, the_year){
  #get proper URL
  the_base_url <- 'https://www.ncwater.org/WUDC/app/LWSP/report.php?'
  the_url <- paste0(the_base_url,'pwsid=',the_facility,'&year=', the_year)
  
  #get the website
  LWSP_website <- read_html(the_url)
  print(the_url)
  
  #scrape the data
  water_system_name <- LWSP_website %>% 
    html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>% 
    html_text()
  PWSID <- LWSP_website %>%  
    html_nodes('td tr:nth-child(1) td:nth-child(5)') %>% 
    html_text()
  Ownership <- LWSP_website %>%  
    html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>%
    html_text()
  MaxDayUse <- LWSP_website %>%  
    html_nodes('th~ td+ td , th~ td+ td') %>%
    html_text()
  
  #convert the the dataframe 
  water_df <- data.frame("Month" = c(1,5,9,2,6,10,3,7,11,4,8,12),
                       "Year" = rep(the_year, 12),
                       "MaxDayUse" = as.numeric(MaxDayUse)) %>% 
            mutate(
              Water_System_Name = !!Ownership,
              PWSID = !!PWSID,
              Ownership = !!Ownership,
              Date = my(paste(Month, "-", Year)))

  #return the dataframe
  return(water_df)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 
durham_water_2020 <- scrape.it('03-32-010',2020)

#graph
ggplot(durham_water_2020, aes(x=Date,
                     y = MaxDayUse))+
  geom_line()+
  labs(
    title = "2020 Durham Muncipal Water Supply Maximum Daily Withdrawal",
    y = "Maximum Daily Withdrawal (mgd)",
    x = "Date")

#show structure
glimpse(durham_water_2020)

```

8. Use the function above to extract data for Cary (PWSID = '03-32-010') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 
#get cary 2020 water data
cary_water_2020 <- scrape.it('03-92-020',2020)

#combine with Durham data
cary_durham_water <- rbind(durham_water_2020, cary_water_2020)

#graph
ggplot(cary_durham_water, aes(x = Date, y = MaxDayUse, color = PWSID))+
  geom_line()+
  labs(
    title = "2020 Durham & Cary Municipal Water Supply Max Daily Withdrawal",
    y = "Maximum Daily Withdrawal (mgd)",
    x = "Date",
    color = "Municipal")+
  scale_color_manual(values = c('green', 'blue'), labels = c('Durham', 'Cary'))

```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 
#create the years and facility vectors
the_years = 2018:2023
the_facility = rep('03-92-020', length(the_years))

#check that they are the same length
the_years
the_facility

#scrape using map2
CaryWater2018to23 <- map2(the_facility, the_years, scrape.it) %>% 
  bind_rows()

#graph it
ggplot(CaryWater2018to2023,aes(x=Date,y=MaxDayUse)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(
    title = "Cary Water Usage Data: 2018-2023",
    y="Max Daily Withdrawal (mgd)",
    x="Date"
    )

```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer:
> Based on the plot there appears to be a strong seasonal trend (increase in the summer and decrease in the winter), and an overall slight increasing trend in Cary water usage over time.   
