---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Sarah Hall"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

**!! AI generated code should NOT be used in this assignment**


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
library(tidyverse)
library(agricolae)
library(lubridate)
library(here)
here()

NTL_LTER <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), stringsAsFactors = TRUE)
#set date columns to date objects
NTL_LTER$sampledate <- mdy(NTL_LTER$sampledate)

#2
#create mytheme, make text size bigger, axis text black, and legend on the top
mytheme <- theme_classic(base_size = 14) 
  theme(axis.text = element_text(color = "black"),
        legend.position = "top")
#set mytheme to standard theme
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperatures in July are the same at all depths
Ha: Mean lake temperatures in July are not the same at all depths (or temp changes with depth) 

4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot, fig.align='left'}
#4
#wrangle data
NTL_LTER_subset <- 
  NTL_LTER %>% 
  select(lakename, year4, daynum, depth, temperature_C) %>% 
  na.omit()

#5
NTL_LTER_subset %>% 
  ggplot(aes(x = depth,
             y = temperature_C))+
  geom_point(alpha = 0.5)+
  geom_smooth(method = "lm",
              se = FALSE,
              col = "blue")+
  labs(
    x = "Depth", 
    y = "Temperature (°C)",
    title = "Temperature change with depth across all lakes in July")+
  ylim(0, 35)
  

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: There appears to be a negative relationship between lake temperature and depth, meaning deeper water tends to be cooler. Points at shallower depths (between 0 to 10) are more widely spread, versus at lower depths the points flatten out. This could suggest that the relationship between temperature and depth is not linear. 


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
TempByDepth.regression <- lm(data = NTL_LTER_subset, temperature_C ~ depth)
summary(TempByDepth.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:
The linear regression shows that there is a strong negative relationship between lake temperature and depth across all lakes in July. The p-value is 2.2e-16, which is less than 0.05 and shows strong statistical significance. The lake temperature is predicted to decreases 1.7°C for every meter of depth. 67.42% of the variance in temperature is explained by the depth, which is based on 34,754 degrees of freedom (the sample size minus 2).  

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
temp_aic <- lm(data = NTL_LTER_subset, temperature_C ~ year4 + daynum + depth)
step(temp_aic)

#10
temp_best.regression <- lm(data = NTL_LTER_subset, temperature_C ~ year4 + 
                             daynum + depth)
summary(temp_best.regression)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method suggest keeping all explanatory variables, so year, day and depth. This model explains 70.26% of the temperature variance, which is a slight improvement over the model using only depth that explained 67.42% of the variance. 

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
#anova test
tempByLake.anova <- aov(data = NTL_LTER_subset, temperature_C ~ lakename)
summary(tempByLake.anova)

#lm 
tempByLake.regression <- lm(data=NTL_LTER_subset, temperature_C ~ lakename)
summary(tempByLake.regression)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: The ANOVA and lm models return a p-value of 2e-16, so we reject the null hypothesis, and conclude that there is a significant difference in mean temperature among the lakes. 


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2, fig.align='left'}
#14.
NTL_LTER_subset %>% 
  ggplot(aes(x = depth,
             y = temperature_C,
             color = lakename))+
  geom_point(alpha = 0.5)+
  geom_smooth(method = "lm",
              se = FALSE,
              aes(color = lakename))+
  labs(
    x = "Depth", 
    y = "Temperature (°C)",
    title = "Temperature change with depth across all lakes in July",
    color = "Lake Name")+
  ylim(0, 35)


```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(tempByLake.anova)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?


>Answer: Ward Lake statistically has the same mean temperature as Peter Lake with a p-value of 0.9997. Central Lake and Crampton Lake have mean temperatures that are statistically distinct from all other lakes.

 
17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: A two-sample t-test looking at if the mean temperature of Peter Lake is significantly different than the mean temperature of Paul Lake.



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
#filter dataset
Crampton_Ward_LakeTTest <- 
  NTL_LTER_subset %>% 
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))
#run ttest
t.test(data = Crampton_Ward_LakeTTest, temperature_C ~ lakename)

```

>Answer: The t-test returns that p-value is 6.264e-07 which is less than 0.05, thus the null hypothesis is rejected. The alternative hypothesis is accepted that the mean temperatures of Crampton Lake and Ward Lake are statistically different. This does match my answer for part 16, where Crampton Lake is one of two lakes that is statistically distinct from all other lakes.  
